Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-07-12 final/co-regular-EPQ.tex
14 pages

%TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.
To define a $\mathsf{MatchedMaxPriority}^>$-complete automaton we need to consider several cases depending on the order between the call/return actions of the $\textit{put}$/$\textit{rm}$ operations that add and respectively, remove the value $b$. For example, the case where the put happens-before the remove (as in \figurename~\ref{fig:introduce gap for EPQ1Lar}) is pictured in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. This automaton captures the three possible ways of ordering the first action $\textit{ret}(\textit{put},a,\_)$ w.r.t. the actions with value $b$, which are pictured in \figurename~\ref{fig:executions APQ1Lar-1 in paper}(a) (this action cannot occur after $\textit{call}(\textit{rm},b,\_)$ since $b$ must be covered by the $a$-s). The paths corresponding to these three possible orders are: $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$, and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$. \figurename~\ref{fig:executions APQ1Lar-1 in paper} lists the four possible orderings of the call/return actions of adding and removing $b$, and also possible orders of the first $\textit{ret}(\textit{put},a,\_)$ w.r.t the actions with value $b$. Each such ordering corresponds to an automaton similar to the one in \figurename~\ref{fig:automata APQ1Lar-1 in paper}, their union defining a $\mathsf{MatchedMaxPriority}^>$-complete automaton.

@@

To characterize violations to linearizability w.r.t. $\mathsf{MatchedMaxPriority}^{=}$ using an automaton that tracks a bounded number of values, we show that the number of values needed to witness that $y <_{\textit{pb}}^* x$, for some $x$ and $y$, is bounded.
%
%While Lemma \ref{lemma:EPQ1Equal as pb order and gap-point} doesn't imply a bound for
%the number of values needs to witness that
%
%The result of Lemma \ref{lemma:EPQ1Equal as pb order and gap-point} still needs to be impove, since obtaining $y <_{\textit{pb}}^* x$ may requires arbitrary many intermediate values, and it is hard to store unbounded values by automata. Fortunately, we prove that the number of intermediate values is essentially bounded, as stated by the following lemma.

@@

To characterize violations to $\mathsf{MatchedMaxPriority}^{=}$-linearizability, one has to consider all the possible orders between call/return actions of the operations on values $a$, $b$, and $a_i$ in Lemma \ref{lemma:ob order has bounded length}, and the right-most gap point of $b$. Excluding the inconsistent cases, we are left with 5 possible orders that are shown in \figurename~\ref{fig:five enumerations}, where $o$ denotes the rightmost gap-point of $b$.
For each case, we define an automaton recognizing the induced set of violations. For instance, the register automaton for the case in \figurename~\ref{fig:five enumerations}(a) is shown in \figurename~\ref{fig:an enumeration and its witness automaton}. In this case, the conditions in Lemma~\ref{lemma:EPQ1Equal as pb order and gap-point} are equivalent to the fact that intuitively, the time interval from $\textit{call}(\textit{rm},a)$ to $\textit{ret}(\textit{rm},b)$ is covered by lower priority values (and thus, there is no gap-point of $b$ which occurs after $\textit{call}(\textit{rm},a)$). By data-independence, these lower priority values can all be renamed to a fixed value $c$. 
%and the other values to a fixed value $\top$. {\color {red}We show how to generate register automata for other cases in the long version \cite{CONCUR2017Ahmed}.}

>>>>

%TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.
To define a $\mathsf{MatchedMaxPriority}^>$-complete automaton, we need to consider all the possible orders between the call/return actions of the $\textit{put}$/$\textit{rm}$ operations that add and respectively, remove the value $b$. The case where the put happens-before the remove (as in \figurename~\ref{fig:introduce gap for EPQ1Lar}) is pictured in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. This automaton captures the three possible ways of ordering the first action $\textit{ret}(\textit{put},a,\_)$ w.r.t. the actions with value $b$, which are pictured in \figurename~\ref{fig:executions APQ1Lar-1 in paper}(a) (this action cannot occur after $\textit{call}(\textit{rm},b,\_)$ since $b$ must be covered by the $a$-s). The paths corresponding to these three possible orders are: $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$, and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$. \figurename~\ref{fig:executions APQ1Lar-1 in paper} lists the four possible orderings of the call/return actions of adding and removing $b$, and also possible orders of the first $\textit{ret}(\textit{put},a,\_)$ w.r.t the actions with value $b$. Each such ordering corresponds to an automaton similar to the one in \figurename~\ref{fig:automata APQ1Lar-1 in paper}, their union defining a $\mathsf{MatchedMaxPriority}^>$-complete automaton.

@@

%To characterize violations to linearizability w.r.t. $\mathsf{MatchedMaxPriority}^{=}$ using an automaton that tracks a bounded number of values, we show 
The following shows that the number of values needed to witness that $y <_{\textit{pb}}^* x$, for some $x$ and $y$, is bounded.
%
%While Lemma \ref{lemma:EPQ1Equal as pb order and gap-point} doesn't imply a bound for
%the number of values needs to witness that
%
%The result of Lemma \ref{lemma:EPQ1Equal as pb order and gap-point} still needs to be impove, since obtaining $y <_{\textit{pb}}^* x$ may requires arbitrary many intermediate values, and it is hard to store unbounded values by automata. Fortunately, we prove that the number of intermediate values is essentially bounded, as stated by the following lemma.

@@

To characterize violations to $\mathsf{MatchedMaxPriority}^{=}$-linearizability, one has to consider all the possible orders between call/return actions of the operations on values $a$, $b$, and $a_i$ in Lemma \ref{lemma:ob order has bounded length}, and the right-most gap point of $b$. Excluding the inconsistent cases, we are left with the five orders in \figurename~\ref{fig:five enumerations}, where $o$ denotes the rightmost gap-point of $b$.
For each case, we define an automaton recognizing the induced set of violations. The register automaton for the case in \figurename~\ref{fig:five enumerations}(a) is shown in \figurename~\ref{fig:an enumeration and its witness automaton}. In this case, Lemma~\ref{lemma:EPQ1Equal as pb order and gap-point} is equivalent to the fact that intuitively, the time interval from $\textit{call}(\textit{rm},a)$ to $\textit{ret}(\textit{rm},b)$ is covered by lower priority values (thus, there is no gap-point of $b$ which occurs after $\textit{call}(\textit{rm},a)$). By data-independence, these lower priority values can be renamed to a fixed value $c$. 
%and the other values to a fixed value $\top$. {\color {red}We show how to generate register automata for other cases in the long version \cite{CONCUR2017Ahmed}.}

### 2017-07-12 final/related.tex
14 pages

Existing automated methods for proving linearizability of a concurrent object
implementation are also based on reductions to safety
verification, e.g.,~\cite{conf/tacas/AbdullaHHJR13, conf/concur/HenzingerSV13,
conf/cav/Vafeiadis10}. The approach in~\cite{conf/cav/Vafeiadis10} considers
implementations where 
operations' \emph{linearization points}
are 
% fixed to particular source-code locations.
manually specified.
Essentially, this approach instruments the
implementation with ghost variables simulating the ADT specification at
linearization points. This approach is incomplete since not all implementations
have fixed linearization points. Aspect-oriented
proofs~\cite{conf/concur/HenzingerSV13} reduce linearizability to the
verification of four simpler safety properties. However, this approach has only
been applied to queues, and has not produced a fully automated
and complete proof technique. The work in~\cite{Dodds:2015:SCT:2676726.2676963} proves 
linearizability of stack implementations with an automated proof assistant. 
Their approach does not lead to full automation however, e.g.,~by reduction to 
safety verification.

>>>>

Existing automated methods for proving linearizability of a concurrent object
 are also based on reductions to safety
verification, e.g.,~\cite{conf/tacas/AbdullaHHJR13, conf/concur/HenzingerSV13,
conf/cav/Vafeiadis10}. The approach in~\cite{conf/cav/Vafeiadis10} considers
implementations where 
operations' \emph{linearization points}
are 
% fixed to particular source-code locations.
manually specified.
Essentially, this approach instruments the
implementation with ghost variables simulating the ADT specification at
linearization points. This approach is incomplete since not all implementations
have fixed linearization points. Aspect-oriented
proofs~\cite{conf/concur/HenzingerSV13} reduce linearizability to the
verification of four simpler safety properties. This approach has only
been applied to queues, and has not produced a fully automated
and complete proof technique. The work in~\cite{Dodds:2015:SCT:2676726.2676963} proves 
linearizability of stack implementations with an automated proof assistant. 
Their approach does not lead to full automation however, e.g.,~by reduction to 
safety verification.
